---
title: "Markdown & Knitr Project 1"
author: "Kerrin Smith"
date: "November 2, 2016"
output: md_document
---

```{r setup, include=FALSE}
## set echo to default to TRUE for all code chunks
knitr::opts_chunk$set(echo = TRUE)
## load ggplot2, gridExtra, and lubridate
library(ggplot2)
library(gridExtra)
library(lubridate)
```

### Loading and preprocessing the data

1. *Load the data (i.e. read.csv())*.
```{r load}
activity <- read.csv("activity.csv")
summary(activity)
```  

2. *Process/transform the data (if necessary) into a format suitable for your
analysis.*  
The dataset appears to be tidy (long, narrow form) and there do not appear to 
be any outliers to address; however, I did choose to convert the date field to 
type Date.

```{r process}
activity$date <- as.Date(activity$date)
```

### What is the mean total number of steps taken per day?

1. *Calculate the total number of steps taken per day.*    

```{r aggrsum}

aggrSum <- aggregate(steps ~ date, data=activity, FUN=sum)
head(aggrSum)

```

2. *Make a histogram of the total number of steps taken each day.*
```{r histStepsPerDay}
hist1 <- qplot(steps, data=aggrSum, binwidth=2500, fill=I("lightgrey"), 
               col=I("black"), main="Histogram of Steps Per Day") +
               theme(panel.background=element_rect(fill="lightblue2"))
print(hist1)
```

3. Calculate and report the mean and median of the total number of steps 
taken per day. **Note: the numbers produced by the summary() function are 
different if the default digits option is used due to rounding. The summary()
output could be corrected using options(digits=10).**

```{r meanMedianTotal}
mean(aggrSum$steps)
median(aggrSum$steps)
```

### What is the average daily activity pattern?

1. *Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis).*

```{r AvgStepsByIntervalPlot}
aggrMean <- aggregate(steps ~ interval, data=activity, FUN=mean)
qplot(interval, steps, data=aggrMean, geom="line", xlab="Interval", 
      ylab="Number of Steps", main="Average Number of Steps by Interval") +
      theme(panel.background=element_rect(fill="lightblue2"))
```

2. *Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?*

```{r MaxMeanInterval}
aggrMean$interval[which.max(aggrMean$steps)]
```


### Imputing missing values
Note that there are a number of days/intervals (in the original data) where
there are missing values (coded as NA). The presence of missing days may
introduce bias into some calculations or summaries of the data.

1. *Calculate and report the total number of missing values in the dataset
(i.e., the total number of rows with NAs).*

```{r totalNAs}
colSums(is.na(activity))
```

2. *Devise a strategy for filling in all of the missing values in the dataset.
The strategy does not need to be sophisticated. For example, you could use
the mean/median for that day, or the mean for that 5-minute interval, etc.*  

The strategy I have chosen is to replace the NAs with the mean for the interval.

3. *Create a new dataset that is equal to the original dataset but with the
missing data filled in.*

```{r newDataset}
activityUpdate <- activity
activityUpdate$steps <- ifelse(is.na(activityUpdate$steps), 
        aggrMean$steps[match(aggrMean$interval, activityUpdate$interval)], 
        activityUpdate$steps)
```

4. *Make a histogram of the total number of steps taken each day and calculate
and report the mean and median total number of steps taken per day. Do these
values differ from the estimates from the first part of the assignment? What is
the impact of imputing missing data on the estimates of the total daily 
number of steps?*  

Imputing the data using the means for the various intervals increases the 
number of days in the mean range. With NAs, there appear to be 19 days in the 
mean range, while the imputed data has 25. The counts for the other ranges 
remain the same. The mean and median are essentially unchanged as well.

```{r histOfImputedData}
aggrSum2 <- aggregate(steps ~ date, data=activityUpdate, FUN=sum)
hist2 <- qplot(steps, data=aggrSum2, binwidth=2500, fill=I("lightgrey"),
               col=I("black"), main="Histogram of Imputed Data") +
               theme(panel.background=element_rect(fill="lightblue2"))
grid.arrange(hist1, hist2, ncol=2)
mean(aggrSum2$steps)
median(aggrSum2$steps)

```

### Are there differences in activity patterns between weekdays and weekends?
1. *Create a new factor variable in the dataset with two levels - "weekday" and
"weekend" indicating whether a given date is a weekday or weekend day.*

```{r addWeekday}
activityUpdate$weekday <- factor(ifelse(wday(activityUpdate$date) %in% c(2:6), 
        "weekday", "weekend"))
table(activityUpdate$weekday)
```

2. *Make a panel plot containing a time series plot (i.e., type="l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis).  

```{r weekdayPanelPlot}
aggrMean2 <- aggregate(steps ~ interval + weekday, 
                       data=activityUpdate, FUN=mean)
ggplot(aggrMean2, aes(x=interval, y=steps)) + xlab("Interval") + 
    ylab("Average Number of Steps") + geom_line() + 
    theme(panel.background=element_rect(fill="lightblue2")) +
    facet_wrap(~weekday, nrow=2) + 
    ggtitle("Activity Patterns: Weekdays vs. Weekends")
```





